---
title: "Sex disparity in liver waitlist outcomes, even if the hazard rate is estimated in an unbiased way"
author: "Hans de Ferrante"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
here::i_am('.here')
pacman::p_load(coxed, ggplot2, tibble, dplyr, stringr, here, purrr, kableExtra, data.table)
options(digits=2)
source(here('settings.R'))
set.seed(1)
```

In this notebook, we illustrate with simulation that sex disparity in liver waitlist 
outcomes can persist, even if prioritization of waitlist candidates is based on a
score which by itself makes predictions of a candidates without-transplant
waitlist mortality without sex bias. 

To this end, we generate realistic populations of liver transplantation candidates
and donors. Without-transplant survival is simulated to depend only on MELD (not on sex).
To allocate the donor livers to patients, we prioritize candidates based on MELD. 
Thus, our prioritization is based on the without-transplant hazard rate faced
by the candidate (which is independent of candidate sex).

In our simulations, however, we simulate females to be on average 15 centimeters
shorter than males, and we simulate short candidates to be less likely to accept
liver offers. We show that this results in substantial sex disparities in waitlist
outcomes (with 5.3% of males and 6.7% of females dead on the waiting list within
90 days of listing, and 5 percentage points fewer females transplanted).

When we fit a Cox proportional hazards model to these data, with waitlist mortality
as the outcome and adjustment for MELD and sex, the sex term is insignificant.
We explain that this is expected, as the Cox proportional hazards model can recover
the true without-transplant survival model in this setting (where no sex bias exists).

The conclusions of this are that (i) using a Cox proportional hazards model
for waitlist mortality is a suitable approach to develop unbiased scores for the
relative without-transplant hazard rate, but (ii) scores developed in such a way 
(e.g., MELD 3.0 and GEMA-Na) cannot correct for disparities which are the result 
of inequities in accessing transplantation.

# Simulation settings

```{r}
# Settings for simulations
NUMBER_YEARS_SIMULATED <- 10
TIME_HORIZON <- NUMBER_YEARS_SIMULATED*365
NUMBER_OF_CANDIDATES_PER_YEAR <- 1700
TOTAL_NUMBER_OF_CANDIDATES <- NUMBER_OF_CANDIDATES_PER_YEAR*NUMBER_YEARS_SIMULATED
```


# Simulation of pre-transplant survival in absence of transplantation

First, we simulate pre-transplant survival for candidates based on a
MELD, and generate candidate sex independently from these survival times.
MELD thus provides an unbiased estimation of waitlist mortality risks for
males and females.

```{r simulate pre-transplant survival}
# Simulate MELD scores according to the empirical distribution of MELD 
# at listing in Eurotransplant
MELD_SCORES <- data.frame(
  meld_lab = sample(
    as.numeric(names(EMPIRICAL_DISTRIBUTION_MELD)), 
    prob=EMPIRICAL_DISTRIBUTION_MELD,
    size=TOTAL_NUMBER_OF_CANDIDATES, 
    replace=TRUE)
)
hist(MELD_SCORES)

# We simulate without-transplant survival based on a Cox proportional hazards model. 
# The marginal effect is set to exp(0.18) - 1. With this marginal effect, the coefficient
# in the Cox proportional hazards model is 0.18. We use a constant hazard function. 
# These parameters were chosen to correspond well to actual waitlist survival 
# by MELD in Eurotransplant.
sim <- sim.survdata(
  N=TOTAL_NUMBER_OF_CANDIDATES, 
  `T`=TIME_HORIZON+1,
  num.data.frames = 1,
  X=MELD_SCORES-18,
  hazard.fun = function(t) {0.0010954},
  beta = exp(0.18)-1
)
simdata <- sim$data
simdata$meld_lab <- simdata$meld_lab + 18

# Plot 90-day survival, estimated with Kaplan-Meier for these MELD scores.
sfit <- survfit(
  Surv(y, failed) ~ meld_lab,
  data=simdata
)

df_surv <- summary(sfit)[
  c('time', 'surv', 'strata', 'lower', 'upper')
  ] |>
  as_tibble() |>
  mutate(
    meld_lab = str_extract(strata, '(?<=meld_lab=).+') |> factor() |> forcats::fct_inorder()
  ) |>
  filter(meld_lab %in% c(6, 15, 20, 25, 30, 35, 40))

ggplot(
  df_surv,
  aes(x = time, y = surv, ymin = lower, ymax = upper, col=meld_lab) 
) +
  geom_step() +
  geom_ribbon(alpha = 0.05) +
  xlim(c(0, 365)) +
  ggokabeito::scale_color_okabe_ito() +
  guides(col = guide_legend('MELD score')) +
  ylab('Pre-transplant survival probability (%)') +
  xlab('Days after registration') +
  theme_bw()
```

Having simulated MELD and the without-transplant survival, we now simulate other patient characteristics.
We point out that these other characteristics are simulated independently of MELD / survival. 
Thus, MELD is still a predictor of a candidate's without-transplant hazard rate without sex bias.


```{r}
simdata$candidate_bloodgroup <- sample(
  factor(names(BLOOD_TYPE_FREQUENCIES)), size = nrow(simdata),
  BLOOD_TYPE_FREQUENCIES, replace=TRUE
)
simdata$candidate_sex <- factor(
  sample(x=c('Male', 'Female'), size=nrow(simdata), prob = c(0.36, 0.64), replace=TRUE),
  levels = c('Male', 'Female')
)
simdata$candidate_height <- if_else(
  simdata$candidate_sex == 'Male',
  rnorm(nrow(simdata), mean = 178, sd = 7),
  rnorm(nrow(simdata), mean = 163, sd = 7)
)

simdata$candidate_country <- sample(factor(1:7), size = nrow(simdata), prob = P_COUNTRY_RATES, replace=TRUE)
summary(simdata)
```
This completes the simulated patient data. For each patient, we have:
  1) a MELD score (`meld_lab`),
  2) a time-to-failure-without-transplant (`y`) which was simulated based on the MELD score,
  3) a death indicator (`failed`), which indicates the candidate would die within 10 years after listing if not transplanted,
  4) a candidate sex (Male or Female),
  5) a candidate height, which depends on sex (Females are on average 15cm shorter),
  6) a candidate blood group
  7) a candidate country of listing

# Simulate a donor population

Secondly, we simulate a donor population. We simulate a donor shortage of 30%.
We assume that donors are uniformly distributed over the 10 years of simulation.
We also randomly draw blood groups and countries for the donor.

```{r}
# Simulate donors. Assume that there is a donor shortage of 30%. Also assume
# that donors are reported uniformly over the simulation period.
NUMBER_OF_DONORS <- 0.70 * TOTAL_NUMBER_OF_CANDIDATES
donor_data = data.frame(
  bloodgroups_donor = sample(
    factor(names(BLOOD_TYPE_FREQUENCIES)),
    size = NUMBER_OF_DONORS,
    BLOOD_TYPE_FREQUENCIES,
    replace=TRUE
  ),
  donor_countries = sample(
    factor(1:7), size = NUMBER_OF_DONORS, prob = D_COUNTRY_RATES, 
    replace=TRUE)
)

summary(donor_data)
```

# Construct a waiting list consisting of all pre-transplant candidates

Here, we construct candidate listing and delisting times in calendar time. 
This means that these listing and delisting times are defined in the number
of days relative to the simulation start time. For instance, a listing time 
of 10 and death time of 140 means that this candidate was activated on the 
waitlist on day 10 of the simulation, and would die 130 (140-10) days after 
listing unless transplanted.

We also sort the complete list of candidates descendingly by MELD. 
This ordering determines the sequence in which candidates are offered
the liver graft, until a candidate accepts the liver graft.

```{r}
# Assume that the initial waitlist size is 70% of the total number
# of candidates reported per year
INITIAL_WAITLIST_SIZE = .70*NUMBER_OF_CANDIDATES_PER_YEAR

# For candidates on the waitlist at simulation start, a registration
# time is generated such that the candidate has not yet died at simulation start.
# For the remaining candidates, we assume that they were registered uniformly
# over the simulation period.
simdata$candidate_listing_time <- if_else(
  runif(nrow(simdata)) < INITIAL_WAITLIST_SIZE/TOTAL_NUMBER_OF_CANDIDATES,
  runif(nrow(simdata), min=-simdata$y, max=0),
  round(runif(nrow(simdata), min = 0, max = TIME_HORIZON), digits=1)
)
simdata$candidate_death_time <- simdata$candidate_listing_time + as.numeric(simdata$y)
simdata$transplanted <- FALSE

# Make the simulation data into a data.table, and order it descendingly
# by MELD.
setDT(simdata)
setorder(simdata, -meld_lab, candidate_listing_time)
simdata$id <- 1:nrow(simdata)

# Print a subset of the simulated data.
kable(arrange(sample_n(simdata, 25), id), 'html') |>
  kable_styling('striped')
```

# Simulate the allocation of donors to candidates.

## Simulate offer acceptance behavior

We simulate offer acceptance behavior based on (i) candidate MELD, and (ii) candidate 
height. For prediction of offer acceptance probabilities, we use logistic 
regressions with unobserved donor heterogeneity. This model was estimated 
based on ET registry data. For the purpose of illustration, we exaggerate 
the effect of candidate height (the real coefficient was about 0.03).

```{r simulate patient offer acceptance probabilies}
# Set up the acceptance formula and fixed effects.
acceptance_model_formula <- ~ I(candidate_height - 170) + meld_lab + f_over_20(meld_lab) + f_over_30(meld_lab)
fixed_effects <- c(
  `(Intercept)` = -2.1,
  `I(candidate_height - 170)` = 0.06,
  `meld_lab` = -0.0146,
  `f_over_20(meld_lab)` = 0.105,
  `f_over_30(meld_lab)` = -0.0753
  )
variance_component_donor_id <- 1.4087

# Prepare transformed X-data for offer acceptance decisions
X <- model.matrix(acceptance_model_formula, data=simdata)[
    , names(fixed_effects)
]
simdata$lp_fixed_effects <- X %*% fixed_effects

# Check the average offer acceptance probability by sex (without a random effect)
simdata %>% split(.$candidate_sex) |> map('lp_fixed_effects') |> map(~ mean(boot::inv.logit(.x)))
```
With this model, females are less likely to accept a graft than males. 
The reason for this is that females are shorter than males (on average 15cm),
which lowers the chance that the candidate accepts a graft offer per our
graft offer acceptance model.

## Start simulation

Here, we iterate over all simulated donors, and place each donor's liver
into a candidate. The sequence in which we offer the liver to candidate's
is based on descending MELD, and graft offer acceptance behavior is simulated
according to the model above.

```{r simulate allocation of donors to patients}
# Assume donors are reported uniformly over the simulation period.
times_donor_reported <- round(
  runif(NUMBER_OF_DONORS, min = 0, max = TIME_HORIZON),
  digits = 2
) |> sort()

# Allow organ offer acceptance decisions to be correlated within each donor.
random_effects_donor_acceptance <- rnorm(
  NUMBER_OF_DONORS, sd = variance_component_donor_id
)

# Simulate the allocation of the donors.
for (i in seq_along(times_donor_reported)) {
  
  # Get the time at which the donor was reported.
  d_reporting_time <- times_donor_reported[[i]]
  d_bloodgroup <- donor_data$bloodgroups_donor[[i]]
  d_country <- donor_data$donor_countries[[i]]
  d_random_effect <- random_effects_donor_acceptance[[i]]
  
  # Eligible candidates have to (i) have a waitlist registration,
  # (ii) not be transplanted already, (iii) listed in the same country as the donor,
  # and (iv) be blood group compatible with the donor
  eligible_candidates <- simdata[
    between(d_reporting_time, candidate_listing_time, candidate_death_time) &
      !transplanted &
      candidate_country == d_country &
      candidate_bloodgroup %in% BLOOD_TYPE_COMPATIBILITIES[[d_bloodgroup]]
  ]
  
  # Acceptance probabilities for patients
  eligible_candidates$prob_accept <- boot::inv.logit(
    eligible_candidates$lp_fixed_effects + d_random_effect
  )
  
  # Simulate acceptance for the candidates. The first candidate simulated
  # to accept the graft gets transplanted.
  u <- runif(n = nrow(eligible_candidates))
  recipient <- eligible_candidates$id[which.min(eligible_candidates$prob_accept <= u)]
  simdata[recipient, c('candidate_transplant_time', 'transplanted') := list(d_reporting_time, TRUE)]
  
  if (i %% 2500 == 0) {
    print(paste0('Simulated ', i, ' placements of liver grafts'))
  }
}
```

## Construct the data we would observe under such liver allocation


```{r}
# Construct the calendar-time at which the candidate exits the waitlist.
# This is the minimum of (i) the maximum time horizon, (ii) the transplant time,
# and (iii) the day the candidate dies on the waiting list.
simdata$candidate_censoring_time <- pmin(
  TIME_HORIZON,
  simdata$candidate_death_time,
  simdata$candidate_transplant_time,
  na.rm=TRUE
  )

# Construct observed outcomes. A waitlist death is observed if the candidate
# was not transplanted in time. The time-to-event, measured from listing, is
# the difference between the censoring time and listing time.
simdata$waitlist_death <- as.integer(!simdata$transplanted & simdata$failed)
simdata$time_to_event <- simdata$candidate_censoring_time - simdata$candidate_listing_time
```
# Assess disparities in waitlist outcomes

Here, we assess sex disparity in waitlist outcomes. To this end, we summarize
by sex the fraction of candidates who (i) is transplanted within 90 days, 
and (ii) died on the waiting list within 90 days.

```{r}
# Prepare a dataset, with time-to-event variables truncated to 90 days.
# This is done commonly when MELD scores are developed.
simdata_truncated <- simdata |>
  mutate(
    transplanted_within_90d = if_else(time_to_event >= 90, 0L, transplanted),
    death_within_90d = if_else(time_to_event >= 90, 0L, waitlist_death),
    time_to_event_within_90d = pmin(time_to_event, 90)
  ) |>
  select(meld_lab, candidate_sex, transplanted_within_90d, death_within_90d, time_to_event_within_90d)

# Show the frequency of waitlist outcomes within 90 days, by sex.
left_join(
  simdata_truncated |>
    group_by(candidate_sex) |>
    summarize(
      `Waitlist deaths within 90 days` = scales::percent(mean(death_within_90d), acc = .1)
    ),
  simdata_truncated |>
    group_by(candidate_sex) |>
    summarize(
      `Transplanted within 90 days` = scales::percent(mean(transplanted_within_90d), acc = .1)
    ),
  by = join_by('candidate_sex')
) |>
  kable('html') |>
  kableExtra::kable_styling('striped')
```

We observe that there is sex disparity in waitlist outcomes;
females are both less likely to have been transplanted within 90 days,
and more likely to have died within 90 days of listing.

# Fitting a Cox proportional hazards model on observed outcomes

Here, we fit a Cox proportional hazards model on the observed data. This model
adjusts for candidate sex and MELD, and candidates without an event 90 days
after listing are censored This approach is commonly used to develop MELD scores,
including MELD 3.0 and GEMA-Na).

```{r}
coxph(
    Surv(time_to_event_within_90d, death_within_90d) ~ meld_lab + candidate_sex,
    data=simdata_truncated
) |>
  broom::tidy(conf.int=TRUE, digits=3) |>
  group_by(row_number(), term) |>
  summarize(
    estimate = glue::glue('{round(estimate, 3)} [{round(conf.low, 3)}-{round(conf.high, 3)}]'),
    p.value=p.value
  ) |>
  mutate(
    `True parameter` = recode(term, meld_lab = 0.18, candidate_sexFemale = 0), .after = term
  ) |>
  select(-1) |>
  kable('html') |>
  kable_styling('striped')
```

Interestingly, the term for female sex is insignificant in this Cox proportional hazards
model, even though more females than males experience a waitlist death.
This is expected: transplantation depended in our simulations only on MELD
and candidate sex. We adjust for these two terms in the waitlist mortality model,
which makes transplantation an independent censoring mechanism. Consequently, the 
Cox model can recover the true model for the without-transplant hazard rate.
In this true model, sex indeed has no effect.

By making short candidates less likely to accept liver offers, we have
simulated a dataset with a sex disparity in waitlist outcomes because
females tend to be shorter than males. However, these disparities will not
show up in the Cox proportional hazards model which uses waitlist mortality
as the outcome. This is because a Cox proportional hazards model does not capture
any excess waitlist mortality which is the result from inequities in access to
transplantation. Other methods are needed to rectify disparities originating
from inequities in access to transplantation.

# Repeating the simulations

The above simulation was conducted only 1 time, and has a point estimate slightly
above 0 for female sex. We can repeat the simulation to confirm that the sex term
is insignificant in general (and not greater than 0):

```{r}
source(here('function_simulate_biased_allocation.R'))
N_ITERATIONS <- 20

sim_results <- map(
  2:N_ITERATIONS %>% setNames(nm = .),
  simulate_biased_allocation
  )

bind_rows(
  sim_results, 
  .id='seed'
  ) |>
  ungroup() |>
  select(-one_of('row_number()')) |>
  filter(str_detect(term, 'sex')) |>
  kable('html') |>
  kable_styling('striped')
```

